\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-25 paper/examples.tex
paper: reworked motivation section

\section{Motivating Example}
Consider an epidemiologist tasked with an 
exploratory analysis of data collected from individuals across
a battery of exams. In particular, she is interested in exploring
the relationship between income and the immune system,
controlling for factors such as weight and gender.

While careful modeling will be necessary before putting
results into the final research paper, the epidemiologist is anxious
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets} for details),
and there is a significant amount of missing data across all
relevant attributes. The researcher must develop a strategy
for handling the missing values. She could drop any records with
missing values and continue her analysis with the remaining data.
She could also fit a model on the base tables, taking advantage of correlations between attributes to fill in the missing values.
However, if she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
Fitting a model on the dataset will take some up-front programming effort and may take a long time to run.

To complicate matters further, the epidemiologist realizes that
some of the steps she is interested in, including filtering and grouping,
are impacted differently by missing data, and that these steps may change
repeatedly, as she wants to run various queries during the exploration phase.
Finally, her preferences may change over time. For example, as she is getting an initial
set of distribution statistics (such as mean and standard deviation), she may be willing to tolerate more error, as long as she gets runtime performance in exchange. Later on,
as more precise estimates are needed for comparing treatments or policy changes, the researcher
may want to focus on the quality of the results.

Rather than work through the implications of each case, she writes her query in standard SQL and submits it to ImputeDB, which automatically performs the imputations necessary to fill in the missing data.
An example query is shown in Fig.~\ref{fig:example-query}.

She indicates her preference for performance or for accuracy by providing a configuration parameter $\alpha$ to ImputeDB in addition to her query.
$\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
So, a high $\alpha$ results in a fast query plan that may compromise accuracy, while a low $\alpha$ results in a high quality query plan that may take longer to run.
In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan} show plans for the example query in Fig.~\ref{fig:example-query} when optimizing for quality and performance, respectively.

In Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan}, $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate.
$\mu$ and $\delta$ are specialized imputation operators.
$\mu$ fills in missing values using a regression model while $\delta$ simply drops tuples which contain a missing value in the specified attributes.
The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
Note that imputation operators are placed in different points of the plan, operating over different columns, and even influencing the order of joins.

The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, rather than dropping tuples with missing values.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan.

@@

\begin{figure}
  \begin{subfigure}[b]{\linewidth}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
  \caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for quality ($\alpha=0$).}
  \label{fig:quality-plan}
\end{subfigure}
\begin{subfigure}[b]{\linewidth}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$}
        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
            [.$\sigma_{\text{exams.weight} \geq 120}$ [.\colorbox{pink}{$\delta_{\text{exams.weight}}$} exams ] ] ] labs ] ] ] ]
\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance ($\alpha=1$).}
\label{fig:fast-plan}
\end{subfigure}
\vspace{0.5\baselineskip}
\caption{Plans for the query in Fig.~\ref{fig:example-query}. Imputation operators are highlighted.}
\end{figure}

Returning to our epidemiologist, tens of queries later she has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.

>>>>

\section{Motivating Example}
In this section, we present a common scenario where a scientists
has to perform analysis on real-world data.

\subsection{Analysis with missing data}
An epidemiologist at the CDC is tasked with an 
exploratory analysis of data collected from individuals across
a battery of exams. In particular, she is interested in exploring
the relationship between income and the immune system,
controlling for factors such as weight and gender.

%While careful modeling will be necessary before putting
%results into the final research paper,

The epidemiologist is excited
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets}),
and there is a significant amount of missing data across all
relevant attributes.

\todo{fill in the XX values}
Before she can perform her queries, the researcher must develop a strategy
for handling the missing values. She currently has two options:
1) she could drop records with missing values in relevant fields,
2) she could use a standard data imputation package on all of her data. Both of these
approaches have significant drawbacks. For the query pictured in 
Fig.~\ref{fig:example-query}, (1) drops XXXX percent of the relevant tuples,
while her previous experiences with (2) have proven be expensive. In a previous project, she
worked with the American Community Survey data (671,153 rows), and performing imputation on the whole
dataset took XXX hours. The analyst needs a more complete picture
of the data, so (1) is insufficient, and she is under time pressure, so (2)
is infeasible.

She can run her queries immediately and finish her report if
she uses \ProjectName{}, which allows her to write her
query in standard SQL and automatically performs the imputations necessary to fill in the missing data.
An example query is shown in Fig.~\ref{fig:example-query}.

@@

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\vspace{0.5\baselineskip}
\caption{A quality-optimized plan for the query in Fig.~\ref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
\label{fig:quality-plan}
\end{figure}

\subsection{Planning with \ProjectName{}}
The multiple possible locations for performing imputation leads to a  search space contains plans of varying performance and quality of results. The user can influence the final plan selected by \ProjectName{}
through a parameter $\alpha \in [0, 1]$, where low $\alpha$ 
prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

For the query in Fig.~\ref{fig:example-query}, \ProjectName{} generates
the plans in Fig.~\ref{fig:quality-plan} and~\ref{fig:fast-plan}, when 
optimizing for quality and performance, respectively.

The quality-optimized plan (Fig.~\ref{fig:quality-plan}) uses the impute operation $\mu$, which employs regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, the performance-optimized plan (Fig.~\ref{fig:fast-plan}) uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.\colorbox{pink}{$\mu_{\text{demo.income}}$}
        [.$\bowtie_{\text{exams.id} = \text{demo.id}}$
          [.\colorbox{pink}{$\mu_{\text{labs.white\_blood\_cell\_ct}}$}
            [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
              [.$\sigma_{\text{exams.weight} \geq 120}$ 
                [.\colorbox{pink}{$\mu_{\text{exams.weight}}$} exams ] 
              ] 
              labs 
            ]
          ]
        [.$\sigma_{\text{demo.gender} = 2}$ demo ]
      ] 
    ] 
  ] 
  ]
\caption{A performance-optimized plan for the query in Fig.~\ref{fig:example-query}. $\sigma$, $\pi$, $\bowtie$, and $g$ are the standard relational selection, projection, join, and group-by/aggregate. $\mu$ and $\delta$ are specialized imputation operators (Sec.~\ref{sec:operators}).}
\label{fig:fast-plan}
\end{figure}

\subsection{\ProjectName{} Workflow}
Now that the analyst has \ProjectName{}, she can explore
the dataset using SQL. For each query, she repeatedly chooses a value
of $\alpha$ and runs the query until she is unsatisfied with the runtime.
Rather than commit \emph{a priori} to a runtime bound for her query,
which may be unsatisfiable, this iterative approach gives her 
immediate feedback.

Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.

%This approach allows her to maximize the query result quality, while
%maintaining an acceptable runtime.
%
%%She indicates her preference for performance or for accuracy by providing a configuration parameter $\alpha$ to ImputeDB in addition to her query.
%%$\alpha$ is a dimensionless parameter in $[0,1]$, and it indicates the relative importance of the quality of query results and of query performance.
%So, a high $\alpha$ results in a fast query plan that may compromise accuracy, while a low $\alpha$ results in a high quality query plan that may take longer to run.
%In our intended workflow, the analyst picks a value of $\alpha$ and incrementally increases it to increase query result quality until the query takes too long to run.
%Fig.~\ref{fig:quality-plan} and Fig.~\ref{fig:fast-plan} show plans for the example query in Fig.~\ref{fig:example-query} when optimizing for quality and performance, respectively.

%1 - How great life is with the project
% - intended work flow

### 2017-02-25 paper/introduction.tex
paper: reworked motivation section

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

>>>>

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution. \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}

\todobox{see how this bit fits}{
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.}
